---
title: "Liver transplant microbiome and MDRB"
author: "Medini K. Annavajhala"
date: "July 19, 2019"
output:
  pdf_document: default
  html_document: default
---
## Generating Figure 4: Overview of findings from multivariate models

This R Markdown document is part of a series used to analyze data and generate figures for the citation below. The current document includes all code and options used to generate the top panel of Figure 4 from the manuscript, which shows the average Shannon $\alpha$--diversity for each of the six major underlying liver disease etiologies at each sampling period.

### Title:
*Colonizing multidrug-resistant bacteria and the longitudinal evolution of the intestinal microbiome after liver transplantation*  

### Authors:
Medini K. Annavajhala, Angela Gomez-Simmonds, Nenad Macesic, Sean B. Sullivan, Anna Kress, Sabrina D. Khan, Marla J. Giddins, Stephania Stump, Grace I. Kim, Ryan Narain, Elizabeth C. Verna, Anne-Catrin Uhlemann  

### Journal: 
*Nature Communications* **2019** 

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=3.5, fig.height=2.5, fig.path='Figure2/', 
                      dev=c('png'), echo=TRUE, warning=FALSE, message=FALSE)
```

### Load Required Libraries: 

```{r libraries}
library("data.table"); packageVersion("data.table")
library("ggplot2"); packageVersion("ggplot2")
library("dplyr"); packageVersion("dplyr")
```

### Import and Format Data

We need to take Shannon diversity values for all samples from patients who had one of the 6 major underlying liver diseases (AIH, ARLD, BILIARY, HBV, HCV, NAFLD) and merge the $\alpha$-diversity by both diagnosis and time category. This gives us an average Shannon index for each time category for each diagnosis, which we can then plot over time

```{r format-alphadt}

alphadt_fig4 = read.table("inputs/Fig4_metadata.txt",sep="\t",header=T)

merged_alphadt_fig4 = aggregate(Shannon~Primary_Diagnosis+Time_Category, 
                                alphadt_fig4, FUN=mean)

```


### Plot and save Figure 4

```{r fig4, fig.width=6, fig.height=2.5, fig.path='Figure4/', dev=c('png'), dpi=300, echo=TRUE, warning=FALSE, message=FALSE}

palette = c("AIH"="#7FC97F","ARLD"="#F0027F","BILIARY"="#BEAED4",
            "HBV"="#FDC086","HCV"="#FFFF99","NAFLD"="#386CB0")

p4 = ggplot(merged_alphadt_fig4,mapping = aes(x=Time_Category, y=Shannon,
                                              color=Primary_Diagnosis, 
                                              group=factor(Primary_Diagnosis))) +
  geom_line(mapping = aes(color=Primary_Diagnosis)) +
  geom_point(size = 1.5) +
  xlab("") + ylab("Shannon Index\n") +
  ggtitle("") +
  scale_color_manual(values=palette) + 
  labs(color="Primary Underlying Etiology") +
  scale_x_discrete(labels= c("Pre-LT","W 1","W 2","W 3",
                             "M 1","M 2","M 3","M 6","M 9","M 12")) + 
  theme(legend.key = element_rect(fill = "white", colour = "white"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        legend.background = element_blank(), panel.background = element_blank(), 
        panel.border = element_rect(fill=NA), axis.line = element_line(colour = "black"), 
        strip.background = element_rect(fill=NA), strip.text = element_text(size=12), 
        legend.title = element_text(size=10), legend.text= element_text(size=8), 
        axis.text = element_text(size=10), axis.title=element_text(size=12),
        legend.position = c(.95, .6), legend.justification = c("right", "top"),
        legend.box.just = "right", legend.margin = margin(1, 1, 1, 1)) + 
  guides(color=guide_legend(ncol=2))

p4
```

The additional components of Figure 4 were drawn and compiled by the author using Inkscape 0.92.2